---
title: "How to use SynTruth"
author: "Dora Schuller"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    theme: united
editor_options: 
  chunk_output_type: console
---

```{r include = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

library(SynTruthPkg)
library(kableExtra)

```


# I. Create a synthetic dataset

Set parameters.

```{r}
# This is an example parameter list with one marker block.

params <- list(
  # parameters for marker blocks
  marker_blocksizes = c(10),
  mus_NB = c(0), 
  mu_diffs = c(2),
  sds_B = c(0.4), 
  sds_NB = c(0.4), 
  rhos_B = c(0.8), 
  rhos_NB = c(0.8), 
  gene_effects = c("AND"),
  
  # parameters for nonmarker blocks 
  mu_nonmarker = 0,
  sds_nonmarker = c(0.4),
  rhos_nonmarker = c(0.8),
  
  # parameters for gep
  n_pts = 1000, 
  n_genes = 100, 
  fraction_pts_benefit = 0.5,
  fraction_tx_1 = 0.5,
  fraction_censored = 0,
  noise = list(type = "random", mean = 0, sd = 1),
  
  # parameters for survival
  surv_distribution = "Weibull",
  surv_d_params = list(shape=1.5),
  scale_NB0 = 10,
  HR_B0_NB0 = 0.7,
  HR_NB1_NB0 = 0.9,
  HR_B1_NB0 = 0.5)

# This is another example parameter list with two marker blocks.

params <- list(
  # parameters for marker blocks
  marker_blocksizes = c(5, 5),
  mus_NB = c(0, 0), 
  mu_diffs = c(1, 2),
  sds_B = c(0.4, 0.4), 
  sds_NB = c(0.4, 0.4), 
  rhos_B = c(0.8, 0.8), 
  rhos_NB = c(0.8, 0.8), 
  gene_effects = c("AND", "no_pattern"),
  
  # parameters for nonmarker blocks 
  mu_nonmarker = 0,
  sds_nonmarker = c(0.3, 0.6),
  rhos_nonmarker = c(0.8, 0.8),
  
  # parameters for gep
  n_pts = 1000, 
  n_genes = 100, 
  fraction_pts_benefit = 0.5,
  fraction_tx_1 = 0.5,
  fraction_censored = 0,
  noise = list(type = "random", mean = 0, sd = 1),
  
  # parameters for survival
  surv_distribution = "Weibull",
  surv_d_params = list(shape=1.5),
  scale_NB0 = 10,
  HR_B0_NB0 = 0.7,
  HR_NB1_NB0 = 0.8,
  HR_B1_NB0 = 0.5)


```


## Run SynTruth once and look into the synthetic data.

```{r}

# This function creates synthetic data and calculates the HRs by fitting a Cox model on patient groups.

gep_hrs <- generate_data(params)

gep <- gep_hrs$gep
hrs <- gep_hrs$hrs
hrs

# Look into GEP:

lookinto <- gep[c(1:2, 501:502), c(1, 100:dim(gep)[2])]
lookinto <- gep[c(495:505), c(1:5, 101:dim(gep)[2])]
is.num <- sapply(lookinto, is.numeric)
lookinto[is.num] <- sapply(lookinto[is.num], round, 2)


lookinto_kable <- kable(lookinto, format='html', 
                           caption="Part of the synthetic dataset") %>% 
  kable_styling(bootstrap_options = "striped", 
                full_width = F, position = "center") 

lookinto_kable

```



## Get the best GEP from n run

Since the HRs estimated by fitting a Cox-model do not always match the user-defined (input) HRs, run SynTruth n_repeats times to get the best GEP, with estimated HRs close to the input.
You have to specify a threshold for the difference of input and output HRs. The first gep, for which the output HRs are lower than this threshold is saved. 

```{r}

n_repeats <- 100
# The get_best_gep function outputs a plot, the best gep and the HRs belonging to the best gep.

gep_n_times <- get_best_gep(params, n_repeats = n_repeats, threshold = 0.03)

# Plot
gep_n_times_plot <- gep_n_times$p
gep_n_times_plot
#save_plot(myfilename = 'cox_pt_group', myplot = cox_pt_group_p, width = 15, height = 10)

```


Check the difference between input and estimated HRs

```{r}

# Extract the best gep:
best_gep <- gep_n_times$best_gep
best_output_hrs <- gep_n_times$best_output_hs
best_output_hrs

# Check difference between input and estimated HRs
diff_io_hrs <- get_diff_io_hrs(input_hrs = c(params$HR_B0_NB0, params$HR_B1_NB0, 
                                             params$HR_NB1_NB0),
                               output_hrs = best_output_hrs,
                               output_format = 'table')

diff_io_hrs_kable <- kable(diff_io_hrs, format='html', 
                           caption="Input and output HRs ") %>% 
  kable_styling(bootstrap_options = "striped", 
                full_width = F, position = "center") 

diff_io_hrs_kable

# Save:
#save_kable(diff_io_hrs_kable, paste("./plots/", myfilename, ".html", sep = ""))

#write.csv(diff_io_hrs, paste('./plots/', myfilename, ".csv", sep = ""))

```



# II. Plot results of best dataset

## Set filename and title

```{r}

myfilename = "example"
mytitle = "Best dataset (example)"

```

## Test for proportional hazards

The plot of scaled Schoenfeld residuals should be a horizontal line. 
The  hypothesis test identifies whether the gradient differs from zero for each variable.

```{r}

test_ph(best_gep)

```


## Get hazard ratio tables

```{r}
# There are two output formats.

hrs <- get_hr_table(best_gep, output_format = 'per_group', mytitle) 
hrs
hrs <- get_hr_table(best_gep, output_format = 'per_benefit', mytitle) 
hrs


# Save:
#myfilename = "example2"
#save_hrs(hrs, myfilename)

```


Plot hazards

```{r}

plot_hazards(best_gep, shape = params$surv_d_params$shape, mytitle = "Hazards")

```


## Plot survival curves

```{r}

survplot <- plot_surv_curve(best_gep, best_output_hrs, mytitle = mytitle)
survplot

#Save survival curves
#myfilename = "example2"
#save_plot(myfilename, myplot = survplot)

```


## Check correlation

```{r}
# Subset the data to check correlation of parts.
marker_genes <- best_gep[,1:sum(params$marker_blocksizes)]
nonmarker_genes <- best_gep[,sum(params$marker_blocksizes):params$n_genes]
whole_gep <- best_gep[,1:params$n_genes]

check_cor(nonmarker_genes, mytitle = mytitle)

```


# III. Explore SynTruth results in scenarios

Does SynTruth construct data according to its settings?

## III/0 Make scenarios

```{r}
# Specify the parameters to be tuned and their values.

scenarios <- list(
  mu_diffs = c(1, 5),
  gene_effects = c('AND', 'OR', "no_pattern")
)

# Specify the rest of the params that stay constant.

params <- list(
  # parameters for marker blocks
  marker_blocksizes = c(10),
  mus_NB = c(0), 
  mu_diffs = c(2),
  sds_B = c(0.4), 
  sds_NB = c(0.4), 
  rhos_B = c(0.8), 
  rhos_NB = c(0.8), 
  gene_effects = c("AND"),
  
  # parameters for nonmarker blocks 
  mu_nonmarker = 0,
  sds_nonmarker = c(0.4),
  rhos_nonmarker = c(0.8),
  
  # parameters for gep
  n_pts = 1000, 
  n_genes = 100, 
  fraction_pts_benefit = 0.5,
  fraction_tx_1 = 0.5,
  fraction_censored = 0,
  noise = list(type = "random", mean = 0, sd = 0),
  
  # parameters for survival
  surv_distribution = "Weibull",
  surv_d_params = list(shape=1.5),
  scale_NB0 = 10,
  HR_B0_NB0 = 0.8,
  HR_NB1_NB0 = 0.7,
  HR_B1_NB0 = 0.5)

# The make_scenarios function outputs:
# 1. a list of parameters, in which each element has a different value of those params that are tuned. This list is the input to functions that run an analysis in different scenarios.
# 2. the names of the scenarios that are composed of the values of the params that are tuned.

scenario_params_names <- make_scenarios(scenarios, params)
scenario_params_list <- scenario_params_names$params_list
scenario_names <- scenario_params_names$sc_names
scenario_names

```

## III/A: Gene expression

### III/A/1: PCA

Is there a separation of B and NB patients?

```{r, fig.height=8, fig.width=14}

scores <- do_pca_in_scenarios(scenario_names, 
                              scenarios = scenario_params_list, 
                              nrow = 2)


#save_plot(myfilename = 'scores', myplot = scores, width = 28, height = 12)


```


### III/A/2 K-means clustering, silhouette score

```{r, fig.height=4, fig.width=7}

km_sc_plot <- do_km_in_scenarios(scenario_names = scenario_names, 
                                 scenarios = scenario_params_list, 
                                 n_repeats = 100,
                                 nrows = 3)
km_sc_plot
#save_plot(myfilename = 'km_sc_plot', myplot = km_sc_plot, width = 14, height = 10)


```

## III/B Can patient groups be predicted from gene expression and treatment?

This part assesses how well the four groups can be reconstructed, by feeding all data – except the true group label - into a hypothetical ‘best’ algorithm. I.e., an algorithm that knows which are the marker genes, and the logical relationship between them. This algorithm is based on the Cox-model.



```{r, fig.height=10, fig.width=15}

cox_pred_hrs <- validate_cox_in_scenarios(scenario_names = scenario_names, 
                                          scenarios = scenario_params_list, 
                                          n_repeats = 100)


cox_pred_hrs

#save_plot(myfilename = 'cox_pred_hrs', myplot = cox_pred_hrs, width = 23, height = 15)
  

```

